---
title: "Fundamentals of R: More Data Structures"
subtitle: "<br><br> STA35A: Statistical Data Science 1"
author: "Xiao Hui Tai"
date: October 2, 2023
output:
  xaringan::moon_reader:
    # css: ["../xaringan-themer.css", "../slides.css"]
    lib_dir: libs
    nature:
      # ratio: "16:9"
      # highlightLines: true
      # highlightStyle: solarized-light
      countIncrementalSlides: false
---

```{r child = "../setup.Rmd"}
```

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
```

```{css, echo = FALSE}
.tiny .remark-code { font-size: 60%; }
.small .remark-code { font-size: 80%; }
```

## Reminders

- Lab is due today at 9pm
  - PDF files only
  - Submission through Gradescope (accessible through Canvas)
  - If you collaborate with others, write their names in your submission

- Office hours: 
  - JH: Tuesdays and Thursdays 10-11 AM MSB 1117
  - XHT: Mondays 2:30-3:30 PM MSB 4242

- First homework will be assigned on Friday; due next Thursday at 9pm

---
## Recap

--

- **Data types**: Logical/Booleans, Integers, Double, Characters, Complex, Raw

- **Operators**: unary, binary, arithmetic, logical, assignment

- **Data structures** group related values together

- **Vectors** group values of the same type

  - Creating and accessing vectors 

---
## Style guide

- https://style.tidyverse.org/index.html


---
## Today
- More on vectors: 
  - Creating vectors 
  - Attributes
  - Vector arithmetic
  - Other functions on vectors
  
- Arrays

---
## Creating vectors
`vector(length = 7)` returns a logical vector of length 7, initialized with FALSEs; helpful for filling things up later (see help page: `vector(mode = "logical", length = 0)`)

```{r}
weeklyHours <- vector(length = 7)
weeklyHours[5] <- 8
weeklyHours
```

The colon operator produces a sequence
```{r}
mySeq <- 2:5
mySeq
```

---
## Creating vectors
Many other ways to produce sequences, e.g., 
```{r}
(mySeq <- seq(from = 1, to = 10, by = 2))
```
<small>(Enclosing an assignment statement in parentheses prints the result)</small>

---
## Vectors with additional attributes: Factors
- **Factors** are built on top of integer vectors

- These have a **fixed and known set of possible values**. 

- Factors have **two components**: level numbers (integers) and level labels (characters) 

```{r}
tmp <- factor(c("BS", "MS", "PhD", "MS"))
tmp
as.integer(tmp)
```

---

## Vectors with additional attributes: Dates

- **Dates and date-times** are built on top of numeric vectors

- Dates are represented internally as the number of days since the origin, 1 Jan 1970

```{r}
z <- as.Date("2020-01-01")
z
typeof(z)
str(z)
```

---

## Vectors with additional attributes: Dates

```{r}
z
as.integer(z)
as.integer(z) / 365 # roughly 50 yrs
```

We will talk more about packages later on in the class, but the `lubridate` package is particularly useful for dealing with dates. 

---
## Vector arithmetic

Operators apply to vectors "pairwise" or "elementwise":
```{r}
x <- c(7, 8, 10, 45)
y <- c(-7, -8, -10, -45)
x + y
x * y
x^c(1, 0, -1, 0.5)
```

---
## Vector arithmetic

R will implicitly coerce the types of vectors to be compatible.  E.g.:
```{r}
TRUE + 4
```
---
## Recycling
- R will also implicitly coerce the length of vectors.

- This is called vector **recycling**: 

  - When a shorter vector is combined with a longer one, elements of the shorter vector are repeated or recycled, to make it the same length as the longer vector.

```{r}
x <- c(7, 8, 10, 45)
x + c(-7, -8)
```

Single numbers are vectors of length 1 for purposes of recycling:
```{r}
2*x
```

---
## Exercises
What do we expect?
```{r eval = FALSE}
myVec <- c(4, 6, 12, 2, 0)
myVec + 1:5
myVec + 1 # recycling
myVec * 5
rep(FALSE, 4) + 1:4 # Coercion
```

---
## Vectorized functions
Most built-in functions are **vectorized**, meaning that they will operate on a vector of numbers:

```{r}
sample(1:10) + 100
```

<small>(what does `sample()` do?)</small>

```{r}
x
x > 9 # pairwise comparisons, where the scalar 9 is recycled
```

---
## Vectorized functions

Lots of functions take vectors as arguments:

- `mean()`, `median()`, `sd()`, `var()`, `max()`, `min()`, `length()`, `sum()`: return single numbers

- `sort()` returns a new vector

- `hist()` takes a vector of numbers and produces a histogram

- `summary()` gives a five-number summary of numerical vectors

- `any()` and `all()` are useful on Boolean vectors   
 

---
## Comparison operators
Boolean operators work elementwise:
```{r}
x
(x > 9) & (x < 20)
```

To get the number of components that satisfy a certain condition:
```{r}
sum(x > 9) # another example of coercion
```


---
## Comparison operators

To compare whole vectors, best to use `identical()` or `all.equal()`:
```{r}
x; y
x == -y
identical(x, -y)
all.equal(x, -y)
```

---
```{r}
identical(c(0.5 - 0.3, 0.3 - 0.1), c(0.3 - 0.1, 0.5 - 0.3))
all.equal(c(0.5 - 0.3, 0.3 - 0.1), c(0.3 - 0.1, 0.5 - 0.3))
```

To see what's going on:
```{r}
print(.5 - .3, digits = 20)
print(.3 - .1, digits = 20)
```

Decimal numbers are not represented exactly in computer arithmetic. For more, see https://stackoverflow.com/questions/9508518/why-are-these-numbers-not-equal.
---
## Combining indexing and comparison operators
```{r}
x; y
x > 9
x[x > 9]
y[x > 9]
```

---
## Combining indexing and comparison operators
`which()` turns a Boolean vector into a vector of TRUE indices:
```{r}
x; y
(places <- which(x > 9))
y[places]
```

---
Note that the behavior of `which()` and the indexing operator is different when there are `NA` values:
```{r}
myNAvec <- c(1, 2, NA)
which(myNAvec > 1)
myNAvec > 1
myNAvec[which(myNAvec > 1)]
myNAvec[myNAvec > 1]
```

---
## Aside: counting the number of missing values
```{r}
myNAvec 
is.na(myNAvec)
sum(is.na(myNAvec))
```

---
## Exercises: comparison operators
What do we expect?
```{r eval = FALSE}
myVec <- c(4, 6, 12, 2, 0)
myVec < 10
(myVec < 10) | (myVec > 1)
myVec + 1 # recycling
myVec * 5
```


---
## Exercises

Get familiar with a few built-in R functions:

```{r eval = FALSE}
myVec <- c(4, 6, 12, 2, 0)
mean(myVec)
median(myVec)
table(myVec)
summary(myVec)
sort(myVec)
```

---
## Named components

You can give names to elements or components of vectors

.small[
```{r}
x
(names(x) <- c("v1", "v2", "v3", "fred"))
x[c("fred", "v1")]
x[c(4, 1)]
```
]

Note the labels are in what R prints; not actually part of the value

---
`names(x)` is just another vector (of characters):
```{r}
names(y) <- names(x)
sort(names(x))
which(names(x) == "fred")
```



---
## Arrays

- Many data structures in R are made by adding bells and whistles to vectors 

- **Arrays** are vectors with *dimensions*

- For example a two-dimensional array:

```{r}
x <- c(7, 8, 10, 45)
x.arr <- array(x, dim = c(2, 2))
x.arr
```

---
## Arrays

- Filled column-wise (by columns)

- `dim` says how many rows and columns

```{r}
dim(x.arr)
```
---

## Arrays with more than two dimensions
- Arrays can have three dimensions ( $r \times c \times h$; think about stacking $r \times c$ matrices)

- Can also have $4, 5, \ldots n$ dimensional arrays

- `dim` is a length $n$ vector
  - Says the size/number of indices of each component
  - e.g., a $4 \times 3$ array has `dim` length 2, elements are 4 and 3

.small[
```{r}
myArr <- array(1:12, dim = c(4, 3))
myArr
dim(myArr)
```
]

---

## Arrays with more than two dimensions
- A $4 \times 3 \times 2$ array has `dim` length 3, elements are 4, 3 and 2

```{r}
myArr <- array(1:24, dim = c(4, 3, 2))
myArr
```

---

## Arrays with more than two dimensions
- A $4 \times 3 \times 2$ array has `dim` length 3, elements are 4, 3 and 2

```{r}
dim(myArr)
```

Some other properties of the array:
```{r}
is.vector(myArr)
is.array(myArr)
```

---
```{r}
typeof(myArr)
str(myArr)
attributes(myArr)
```
`typeof()` returns the type of the _elements_

`str()` gives the **structure**: here, a numeric array, with three dimensions, size/indices, and then the actual numbers


---
## Summary

--
- Vectors: 
  - Additional attributes: Factors, dates
  - Vector arithmetic
  - Other functions on vectors
    - Comparison operators 
    - Indexing operators
  - Named components

- Arrays
  - Vectors with dimension
  
