---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

```{r setup, message=FALSE, warning=FALSE, echo=FALSE}
################################################################################
# DEFAULT SETTINGS
################################################################################
knitr::opts_chunk$set(message = FALSE) # don't knit messages
knitr::opts_chunk$set(warning = FALSE) # don't knit warnings
```

# Website for STA 35A at UC Davis, Fall 2022

## Attribution
Some of the materials are adapted from [Data Science in a Box](https://datasciencebox.org/) by Mine Çetinkaya-Rundel, [STA 199](https://sta199-fa21-003.netlify.app/) by Maria Tackett, [STA 198](https://sta198f2021.github.io/website/) by Amy Herring, and [Statistical Computing 36-350: Beginning to Advanced Techniques in R](http://www.stat.cmu.edu/~cshalizi/statcomp/14 ) by Shalizi, C. R. and Thomas, A. C..

## License
This work is licensed under a [Creative Commons Attribution-NonCommercial 4.0 International (CC BY-NC 4.0) License](https://creativecommons.org/licenses/by-nc/4.0/).
