---
title: "Data visualization and Descriptive Statistics"
subtitle: "<br><br> STA35A: Statistical Data Science 1"
author: "Xiao Hui Tai"
date: October 10, 2022
output:
  xaringan::moon_reader:
    # css: ["../xaringan-themer.css", "../slides.css"]
    lib_dir: libs
    nature:
      # ratio: "16:9"
      # highlightLines: true
      # highlightStyle: solarized-light
      countIncrementalSlides: false
---

```{r child = "../setup.Rmd"}
```

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(openintro)
library(palmerpenguins)
```

```{css, echo = FALSE}
.tiny .remark-code { font-size: 60%; }
.small .remark-code { font-size: 80%; }
.tiny { font-size: 60%; }
.small { font-size: 80%; }
```

## Reminders/announcements

- Lab 3 is due today at 9pm

- Participation survey closes at 9pm

- HW 2 posted on course website, due Thursday at 9pm

- Schedule for next week (week 5):
  - Monday: 9:30-10:30 OH (XHT); regular lecture; lab 4 due 9pm, participation survey closes 9pm
  - Tuesday: 10-11 OH (JH)
  - Wednesday: Jed Harwood will do review during regular lecture time (same room)
  - Thursday: 10:30-11:30 OH (JH); **no lab**, instead 12:10-2PM OH (XHT; virtual, will post link on Piazza)
  - Friday: midterm during regular lecture time (same room); **no homework**

---
## Midterm
- Midterm will cover material until Monday, Oct 17

- Closed-book

- You don't need your computers

- There will be no make-up exams

- Drop policy for exams: 1 midterm may be dropped or used with 50% penalty on the final 

- Updated syllabus: "There will be no make-up exams. If you must miss an exam due to illness, travel, or some other reason, this will be the exam that will be dropped. For the final, if you have another final starting 30 minutes before or after the scheduled time, you may present documentation and request for an accommodation to start 15 minutes before or after the scheduled time."

- Exam dates were announced on the first day of class: Oct 21, Nov 18 and Dec 7

---
## Recap
--

- More on `ggplot()` 
  - Data, mapping, geom, labels 

  - Mapping vs. setting
  
  - Faceting using `facet_wrap()` and `facet_grid()`

---
## Questions from last time 

- Centering title: `+ theme(plot.title = element_text(hjust = 0.5))`

- Change number of breaks in legend: `+ scale_size_continuous(breaks = seq(2, 14, by = 2))`, or `scale_alpha_continuous()`, ... 

- Setting `size` twice does not print a warning; just overrides the first 

.small[
.pull-left[
```{r eval = FALSE}
ggplot(penguins,
       aes(x = bill_depth_mm,
           y = bill_length_mm,
           size = body_mass_g)) + 
  geom_point(size = 2, alpha = 0.5) 
```
]

.pull-left[
```{r echo = FALSE, out.width = "100%"}
ggplot(penguins,
       aes(x = bill_depth_mm,
           y = bill_length_mm,
           size = body_mass_g)) + 
  geom_point(size = 2, alpha = 0.5) 
```
]
]


---
## Today

- Finishing up on `ggplot()`

  - Faceting using `facet_grid()`

  - Misc. items
  
- Descriptive statistics 

---
## 2D grid using `facet_grid()`

- `facet_grid()`:
    - 2D grid
    - `rows ~ cols`
    - use `.` for no split (1D)

- Uses all levels, even if there are no observations; i.e., may produce empty plots 

- `facet_wrap()` only displays the plots having actual values
---

## 2D grid using `facet_grid()`

```{r warning = FALSE}
ggplot(penguins, aes(x = bill_depth_mm, y = bill_length_mm)) + 
  geom_point() +
  facet_grid(species ~ sex) #<<
```

---
## 2D grid using `facet_grid()`

```{r warning = FALSE}
ggplot(penguins, aes(x = bill_depth_mm, y = bill_length_mm)) + 
  geom_point() +
  facet_grid(sex ~ species) #<<
```

---
## 1D grid using `facet_grid()`

```{r warning = FALSE, fig.asp = 0.5}
ggplot(penguins, aes(x = bill_depth_mm, y = bill_length_mm)) + 
  geom_point() +
  facet_grid(. ~ species) #<<
```

---
## 1D grid using `facet_grid()`

```{r warning = FALSE}
ggplot(penguins, aes(x = bill_depth_mm, y = bill_length_mm)) + 
  geom_point() +
  facet_grid(species ~ .) #<<
```

---

## Facet and color

.pull-left[
```{r facet-color-legend, fig.show = "hide", warning = FALSE}
ggplot(
  penguins, 
  aes(x = bill_depth_mm, 
      y = bill_length_mm, 
      color = species)) + #<<
  geom_point() +
  facet_grid(species ~ sex) +
  scale_color_viridis_d() #<<
```
]

.pull-right[
```{r ref.label = "facet-color-legend", echo = FALSE, warning = FALSE, out.width = "100%"}
```
]

---

## Facet and color, no legend

.pull-left[
```{r facet-color-no-legend, fig.show = "hide", warning = FALSE}
ggplot(
  penguins, 
  aes(x = bill_depth_mm, 
      y = bill_length_mm, 
      color = species)) +
  geom_point() +
  facet_grid(species ~ sex) +
  scale_color_viridis_d() +
  guides(color = "none") #<<
```
]

.pull-right[
```{r ref.label = "facet-color-no-legend", echo = FALSE, warning = FALSE, out.width = "100%"}
```
]

---
## A note on piping and layering

- Pipe `%>%` used mainly in `dplyr` pipelines
  - Pipe the output of the previous line of code as the first input of the next line of code

- `+` used in `ggplot2` plots is used for "layering"
  - Create the plot in layers, separated by `+`

---
```{r message=FALSE, echo = FALSE}
hotels <- readr::read_csv("../lecture6_manipulation2/data/hotels.csv")
```

## dplyr

`r emo::ji("x")`

```{r error=TRUE}
hotels +
  select(hotel, lead_time)
```

`r emo::ji("white_check_mark")`

```{r eval=FALSE}
hotels %>%
  select(hotel, lead_time)
```

.tiny[
```{r echo=FALSE, output.lines=6}
hotels %>%
  select(hotel, lead_time)
```
]

---

## ggplot2

`r emo::ji("x")`

.small[
```{r error=TRUE}
ggplot(hotels, aes(x = hotel, fill = deposit_type)) %>%
  geom_bar()
```
]

`r emo::ji("white_check_mark")`

```{r out.width="25%"}
ggplot(hotels, aes(x = hotel, fill = deposit_type)) +
  geom_bar()
```


---
## Code styling

Many of the styling principles are consistent across `%>%` and `+`:

- always a space before
- always a line break after (for pipelines with more than 2 lines)

`r emo::ji("x")`

```{r eval=FALSE}
ggplot(hotels,aes(x=hotel,y=deposit_type))+geom_bar()
```

`r emo::ji("white_check_mark")`

```{r eval=FALSE}
ggplot(hotels, aes(x = hotel, y = deposit_type)) + 
  geom_bar()
```

---
## Back to UN Votes visualization from first lecture 


```{r echo = FALSE, out.width = '100%'}
library(unvotes)
un_votes %>%
  filter(country %in% c("United States", "United Kingdom", "China", "Singapore")) %>%
  inner_join(un_roll_calls, by = "rcid") %>%
  inner_join(un_roll_call_issues, by = "rcid") %>%
  mutate(year = lubridate::year(date)) %>%
  group_by(country, year, issue) %>%
  summarize(votes = n(),
            percent_yes = mean(vote == "yes")) %>%
  filter(votes > 5) %>%  # Only use records where there are more than 5 votes
  ggplot(mapping = aes(x = year, y = percent_yes, color = country)) +
    geom_point(alpha = 0.4) +
    geom_smooth(method = "loess", se = FALSE) +
    facet_wrap(~ issue) +
    scale_y_continuous(labels = scales::percent) +
    labs(
      title = "Percentage of 'Yes' votes in the UN General Assembly",
      subtitle = "1946 to 2019",
      y = "% Yes",
      x = "Year",
      color = "Country"
    ) +
    scale_color_viridis_d() +
    theme(text = element_text(size = 9))   
```

---
<br><br>
.small[
```{r eval = FALSE}
un_votes %>%
  filter(country %in% c("United States", "United Kingdom", "China", "Singapore")) %>%
  inner_join(un_roll_calls, by = "rcid") %>%
  inner_join(un_roll_call_issues, by = "rcid") %>%
  mutate(year = lubridate::year(date)) %>%
  group_by(country, year, issue) %>%
  summarize(votes = n(),
            percent_yes = mean(vote == "yes")) %>%
  filter(votes > 5) %>%  # Only use records where there are more than 5 votes
  ggplot(mapping = aes(x = year, y = percent_yes, color = country)) +
    geom_point(alpha = 0.4) +
    geom_smooth(method = "loess", se = FALSE) +
    facet_wrap(~issue) +
    scale_y_continuous(labels = scales::percent) +
    labs(
      title = "Percentage of 'Yes' votes in the UN General Assembly",
      subtitle = "1946 to 2019",
      y = "% Yes",
      x = "Year",
      color = "Country"
    ) +
    scale_color_viridis_d() +
    theme(text = element_text(size = 9))   
```
]

---
## Course content 

1. Fundamentals of R
  - Overview of data types and structures
  - Data manipulation and data visualization tools  

2. **Descriptive statistics for numerical and categorical data**

3. Probability
  - Rules of probability computation; conditional probability
  - Basic probability models: Binomial, Normal and Poisson 

4. Statistical inference
  - Sampling distributions of sample mean and sample proportion 
  - Hypothesis testing and confidence intervals for population mean and population proportion

---
## Descriptive statistics

- We've now learned about data manipulation and visualization tools

- What visualizations to do and what summary statistics to actually calculate?

- **Descriptive statistics** are numbers that are used to summarize and describe data

- **Numerical** or **graphical** ways to display the data

- Why is this a useful thing to do?

**Ages of students**: 18, 19, 20, 18, 20, 20, 19, 18, 19, 20, 18, 20, 20, 19, 18, 19, 20, 18, 20, 20, 19, 18, 19, 20, 18, 20, 20, 19, 18, 19, 20, 18, 20, 20, 19, 18, 19, 20, 18, 20, 20, 19, 18, 19, 20, 18, 20, 20, 19, 18, 19, 20, 18, 20, 20, 19, 18, 19, 20, 18, 20, 20, 19, 18, 19, 20, 18, 20, 20, 19, 18, 19, 20, 18, 20, 20, 19, 18, 19, 20, 18, 20, 20, 19, 18, 19, 20, 18, 20, 20, 19

---

## Terminology: Number of variables involved

- **Univariate** data analysis: distribution of single variable

- **Bivariate** data analysis: relationship between two variables

- **Multivariate** data analysis: relationship between many variables at once, usually focusing on the relationship between two while conditioning for others

---

## Terminology: Types of variables

- **Numerical** variables
  - E.g., age, length, temperature
  
  - **Continuous** variables can take on an infinite number of values
  
  - **Discrete** variables only take on non-negative whole numbers

- **Categorical** variables

  - E.g., year in college, type of bike, meal 
  
  - **Ordinal** variables have levels that have a natural ordering

---
## Data: Lending Club

- Lending Club is a platform that allows individuals to lend to other individuals

- Data are available in the `openintro` package, called `loans_full_schema`

- Includes 10,000 loans made through the Lending Club; has 55 columns

.tiny[
```{r }
library(openintro)
dplyr::glimpse(loans_full_schema) 
```
]
---

## Selected variables

```{r}
loans <- loans_full_schema %>%
  select(loan_amount, interest_rate, term, grade, 
         state, annual_income, homeownership, debt_to_income,
         issue_month)
glimpse(loans)
```

---

## Selected variables

.small[
Variable        | Description
----------------|-------------
`loan_amount`   |	Amount of the loan received, in US dollars
`interest_rate` |	Interest rate on the loan, in an annual percentage
`term`	        | The length of the loan, which is always set as a whole number of months
`grade`	        | Loan grade, which takes a values A through G and represents the quality of the loan and its likelihood of being repaid
`state`         |	US state where the borrower resides
`annual_income` |	Borrower’s annual income, including any second income, in US dollars
`homeownership`	| Indicates whether the person owns, owns but has a mortgage, or rents
`debt_to_income` | Debt-to-income ratio
`issue_month` | Month the loan was issued
]

---

## Variable types

.small[
Variable        | Description
----------------|-------------
`loan_amount`   |	Amount of the loan received, in US dollars
`interest_rate` |	Interest rate on the loan, in an annual percentage
`term`	        | The length of the loan, which is always set as a whole number of months
`grade`	        | Loan grade, which takes a values A through G and represents the quality of the loan and its likelihood of being repaid
`state`         |	US state where the borrower resides
`annual_income` |	Borrower’s annual income, including any second income, in US dollars
`homeownership`	| Indicates whether the person owns, owns but has a mortgage, or rents
`debt_to_income` | Debt-to-income ratio
`issue_month` | Month the loan was issued
]

- Numerical variables: Continuous or discrete?
- Categorical: Ordinal or not? 

---

## Variable types

Variable        | Type
----------------|-------------
`loan_amount`   |	numerical, continuous
`interest_rate` |	numerical, continuous
`term`	        | numerical, discrete
`grade`	        | categorical, ordinal
`state`         |	categorical, not ordinal
`annual_income` |	numerical, continuous
`homeownership`	| categorical, not ordinal
`debt_to_income` | numerical, continuous
`issue_month` | date 


---

## Describing numerical distributions

- **Visual summaries**:
  - Histogram
  - Boxplot
  - Density plot
  - Line graph 
  
- Measures of **central tendency**: mean, median, mode 

- **Shape**:
    - Skewness: right-skewed, left-skewed, symmetric 
    - Modality: unimodal, bimodal, multimodal, uniform

- **Spread**: variance and standard deviation, range and interquartile range

- **Unusual observations**

- A **summary statistic** is a single number summarizing a large amount of data

---

## Histogram

- Shows **shape, center, and spread** of the data

- Contiguous (adjoining) boxes
  - Horizontal axis: what the data represents
  - Vertical axis: frequency or relative frequency 

.tiny[
.pull-left[
```{r message = TRUE, out.width = "100%"}
ggplot(loans, aes(x = loan_amount)) +
  geom_histogram()
```
]

.pull-right[
```{r out.width = "100%"}
hist(loans_full_schema$loan_amount)
```
]
]

---
## Histograms and binwidth

.panelset[
.panel[.panel-name[binwidth = 1000]
```{r out.width = "50%"}
ggplot(loans, aes(x = loan_amount)) +
  geom_histogram(binwidth = 1000)
```
]
.panel[.panel-name[binwidth = 5000]
```{r out.width = "50%"}
ggplot(loans, aes(x = loan_amount)) +
  geom_histogram(binwidth = 5000)
```
]
.panel[.panel-name[binwidth = 20000]
```{r out.width = "50%"}
ggplot(loans, aes(x = loan_amount)) +
  geom_histogram(binwidth = 20000)
```
]
]

---

## Adding labels 

.panelset[
.panel[.panel-name[Plot]
```{r ref.label = "hist-custom", echo = FALSE, warning = FALSE}
```
]
.panel[.panel-name[Code]
```{r hist-custom, fig.show = "hide", warning = FALSE}
ggplot(loans, aes(x = loan_amount)) +
  geom_histogram(binwidth = 5000) +
  labs( #<<
    x = "Loan amount ($)", #<<
    y = "Frequency", #<<
    title = "Amounts of Lending Club loans" #<<
  ) #<<
```
]
]


---
## Summary
--

- Descriptive statistics

  - Types of variables (numerical and categorical)
  
- Describing numerical distributions

  - Histograms
  
  


