---
title: "Descriptive Statistics: Numerical and Categorical Data"
subtitle: "<br><br> STA35A: Statistical Data Science 1"
author: "Xiao Hui Tai"
date: October 17, 2022
output:
  xaringan::moon_reader:
    # css: ["../xaringan-themer.css", "../slides.css"]
    lib_dir: libs
    nature:
      # ratio: "16:9"
      # highlightLines: true
      # highlightStyle: solarized-light
      countIncrementalSlides: false
---

```{r child = "../setup.Rmd"}
```

```{css, echo = FALSE}
.tiny .remark-code { font-size: 60%; }
.small .remark-code { font-size: 80%; }
```


```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(openintro)
```

## Recap
- Relationships between numerical variables 

  - Correlation coefficient
  
  - Line graph

- Describing categorical distributions

  - Bar plot

- Relationships between categorical data

  - Contingency tables
  
---
## Today

- Relationships between categorical data

  - Contingency tables
  
  - Stacked bar plot

- Relationships between numerical and categorical data 

  - Fill and facet
  
  - Side-by-side boxplots
  
  - Other fancy plots

  - Numerical summaries in R

---
## Data: Lending Club

- Lending Club is a platform that allows individuals to lend to other individuals

```{r}
loans <- loans_full_schema %>%
  select(loan_amount, interest_rate, term, grade, 
         state, annual_income, homeownership, debt_to_income,
         issue_month)
glimpse(loans)
```

---
## Contingency tables
A contingency table summarizes data for two categorical variables

```{r}
xtabs(~ homeownership + grade, data = loans_full_schema)
```

Each value in the table represents the number of times a particular combination of variable outcomes occurred, in other words the **frequency distribution** of the variables

---
## Contingency tables with row and column totals
```{r}
outTable <- xtabs(~ homeownership + grade, data = loans_full_schema)
outTableTotals <- outTable %>%
  cbind(rowTotal = rowSums(outTable)) 
outTableTotals <- outTableTotals %>%
  rbind(columnTotal = colSums(outTableTotals))
outTableTotals
```


---
## Contingency tables with proportions
- How do we code **row proportions** in R?

.tiny[
```{r}
prop.table(outTable, margin = 1)
```
]

- Note that each row should sum to 1

---
## Contingency tables with proportions
- `dplyr` option is not so straightforward

.tiny[
```{r}
outTableTotals %>%
  data.frame() %>%
  mutate(across(everything(), ~ .x/rowTotal))
```
]
---
## Contingency tables with proportions
- `apply` option is also not straightforward

.tiny[
```{r}
t(apply(outTableTotals, MARGIN = 1, FUN = function(x) x/x[9]))
```
]

---
## Contingency tables with proportions
- How about **column proportions**?


---
## Contingency tables with proportions
- How about **column proportions**?

.tiny[
```{r}
prop.table(outTable, margin = 2)
```
]

- Note that each column should sum to 1

---
## Stacked bar plot

- A stacked bar plot looks at numeric values across two categorical variable

- Each bar in a standard bar plot is divided into stacked sub-bars, each one corresponding to a level of the second categorical variable.

```{r}
ggplot(loans, aes(x = homeownership, 
                  fill = grade)) + #<<
  geom_bar()
```
---

## Stacked bar plot
Turning `grade` into an ordered variable makes `ggplot` use the `viridis` scale by default

```{r}
str(loans$grade)
loans <- loans %>%
  mutate(grade = factor(grade, ordered = TRUE))
str(loans$grade)
```

---
## Stacked bar plot
```{r}
ggplot(loans, aes(x = homeownership, 
                  fill = grade)) + #<<
  geom_bar()
```

---

## Stacked bar plot
Adding `position = "fill"` argument changes visualization to proportions, and standardizes the height of columns

```{r}
ggplot(loans, aes(x = homeownership, fill = grade)) +
  geom_bar(position = "fill") #<<
```

---
## Stacked bar plot: counts vs. proportions
Which bar plot is a more useful representation for visualizing the relationship between homeownership and grade?


.pull-left[
```{r echo=FALSE, out.width = "100%"}
ggplot(loans, aes(x = homeownership, fill = grade)) +
  geom_bar()
```
]
.pull-right[
```{r echo=FALSE, out.width = "100%"}
ggplot(loans, aes(x = homeownership, fill = grade)) +
  geom_bar(position = "fill")
```
]

If there were no relationship between homeownership and grade, we would expect to see the bars to be similar lengths across the homeownership status (columns).

---
## Stacked bar plot: counts vs. proportions
Is there a relationship between homeownership and grade? 

.pull-left[
```{r echo=FALSE, out.width = "100%"}
ggplot(loans, aes(x = homeownership, fill = grade)) +
  geom_bar()
```
]
.pull-right[
```{r echo=FALSE, out.width = "100%"}
ggplot(loans, aes(x = homeownership, fill = grade)) +
  geom_bar(position = "fill")
```
]


---

## Customizing bar plots

.panelset[
.panel[.panel-name[Plot]
```{r ref.label = "bar-custom", echo = FALSE, warning = FALSE}
```
]
.panel[.panel-name[Code]
```{r bar-custom, fig.show = "hide", warning = FALSE}
ggplot(loans, aes(y = homeownership, #<<
                  fill = grade)) +
  geom_bar(position = "fill") +
  labs( #<<
    x = "Proportion", #<<
    y = "Homeownership", #<<
    fill = "Grade", #<<
    title = "Grades of Lending Club loans", #<<
    subtitle = "and homeownership of lendee" #<<
  ) #<<
```
]
]

---
## Relationships between numerical and categorical data 
- We saw histograms, boxplots, and density plots earlier, for describing a single numerical variable

- To look at relationships between these numerical data and a categorical variable, we can:

  - Fill and facet histograms and density plots
  
  - Use side-by-side boxplots
  
  - Violin plots
  
  - Ridge plots

- Numerical summaries
  - `group_by()`

---

## Fill a histogram with a categorical variable

Is there a relationship between loan amount and home-ownership status? 

.panelset[
.panel[.panel-name[Plot]
```{r ref.label = "hist-fill", echo = FALSE, warning = FALSE}
```
]
.panel[.panel-name[Code]
```{r hist-fill, fig.show = "hide", warning = FALSE}
ggplot(loans, aes(x = loan_amount, 
                  fill = homeownership)) + #<<
  geom_histogram(binwidth = 5000,
                 alpha = 0.5) + #<<
  labs(
    x = "Loan amount ($)",
    y = "Frequency",
    title = "Amounts of Lending Club loans"
  ) +
  scale_fill_viridis_d()
```
]
]

---

## Fill a histogram with a categorical variable

Is there a relationship between loan amount and home-ownership status? 
  - Need `position = "identity"` argument if we don't want histogram bars to be stacked on top of one another 

.panelset[
.panel[.panel-name[Plot]
```{r ref.label = "hist-fill2", echo = FALSE, warning = FALSE}
```
]
.panel[.panel-name[Code]
```{r hist-fill2, fig.show = "hide", warning = FALSE}
ggplot(loans, aes(x = loan_amount, 
                  fill = homeownership)) + 
  geom_histogram(position = "identity", #<<
                 binwidth = 5000,
                 alpha = 0.5) + 
  labs(
    x = "Loan amount ($)",
    y = "Frequency",
    title = "Amounts of Lending Club loans"
  ) +
  scale_fill_viridis_d()
```
]
]


---

## Facet a histogram with a categorical variable

Is there a relationship between loan amount and home-ownership status? 

.panelset[
.panel[.panel-name[Plot]
```{r ref.label = "hist-facet", echo = FALSE, warning = FALSE}
```
]
.panel[.panel-name[Code]
```{r hist-facet, fig.show = "hide", warning = FALSE}
ggplot(loans, aes(x = loan_amount, fill = homeownership)) + 
  geom_histogram(binwidth = 5000) +
  labs(
    x = "Loan amount ($)",
    y = "Frequency",
    title = "Amounts of Lending Club loans"
  ) +
  facet_wrap(~ homeownership, nrow = 3) + #<<
  scale_fill_viridis_d()
```
]
]

---
## Filling density plots with a categorical variable 

Is there a relationship between loan amount and home-ownership status? 

.panelset[
.panel[.panel-name[Plot]
```{r ref.label = "density-cat", echo = FALSE, warning = FALSE}
```
]
.panel[.panel-name[Code]
```{r density-cat, fig.show = "hide", warning = FALSE}
ggplot(loans, aes(x = loan_amount, 
                  fill = homeownership)) + #<<
  geom_density(adjust = 2, 
               alpha = 0.5) + #<<
  labs(
    x = "Loan amount ($)",
    y = "Density",
    title = "Amounts of Lending Club loans", 
    fill = "Homeownership" #<<
  ) +
  scale_fill_viridis_d()
```
]
]


---

## Side-by-side boxplots

Is there a relationship between loan amount and home-ownership status? 

.panelset[
.panel[.panel-name[Plot]
```{r ref.label = "box-cat", echo = FALSE, warning = FALSE}
```
]
.panel[.panel-name[Code]
```{r box-cat, fig.show = "hide", warning = FALSE}
ggplot(loans, aes(x = loan_amount,
                  y = homeownership)) + #<<
  geom_boxplot() +
  labs(
    x = "Loan amount ($)",
    y = "Home-ownership status",
    title = "Amounts of Lending Club loans" #<<
  )
```
]
]

---

## Violin plots

Is there a relationship between loan amount and home-ownership status? 

- A violin plot is a hybrid of a boxplot and a density plot

```{r warning = FALSE}
ggplot(loans, aes(x = homeownership, y = loan_amount)) +
  geom_violin()
```


---

## Ridge plots

```{r warning = FALSE}
library(ggridges)
ggplot(loans, aes(x = loan_amount, y = homeownership, 
                  fill = homeownership)) + 
  geom_density_ridges(alpha = 0.5) +
  scale_fill_viridis_d()
```

---

## Ridge plots

Ridge plots can also be used to investigate more complicated relationships, such as those between a categorical and numerical variable, conditional on another categorical variable 

Here, we consider the relationship between loan grade and loan amount, conditional on each level of home ownership

```{r warning = FALSE}
ggplot(loans, aes(x = loan_amount, y = homeownership, fill = grade, color = grade)) + 
  geom_density_ridges(alpha = 0.5)
```


---

## Ridge plots

Here, we consider the relationship between loan grade and loan amount, conditional on each level of home ownership

```{r warning = FALSE}
ggplot(loans, aes(x = loan_amount, y = homeownership, fill = grade, color = grade)) + 
  geom_density_ridges(alpha = 0.5)
```

Interestingly, those who had mortgages tend to have a higher proportion of grade G loans that have higher loan amounts. 

---
## Numerical summaries in R, grouping by a categorical variable 
Question: `homeownership` is a factor variable with three levels, `MORTGAGE`, `OWN` and `RENT`. How do we calculate the mean loan amount for each type of home ownership status? 

--
```{r}
loans %>%
  group_by(homeownership) %>%
  summarize(meanLoan = mean(loan_amount))
```

---

## What descriptive statistics to use? Which plots to produce?

- Fancier does not always mean better; a pretty plot can look great but tell us nothing 

- Think about what question you are trying to answer and pick the figure that best suits the purpose

- Hadley Wickham on exploratory data analysis: "EDA is not a formal process with a strict set of rules. More than anything, EDA is a state of mind." (Chapter 7, R for Data Science)



---
## Summary

--
- Relationships between categorical data

  - Contingency tables
  
  - Stacked bar plot

- Relationships between numerical and categorical data 
  - Fill and facet
  
  - Side-by-side boxplots
  
  - Other fancy plots
  
  - Numerical summaries in R

