---
title: "Poisson Distribution"
subtitle: "<br><br> STA35A: Statistical Data Science 1"
author: "Xiao Hui Tai"
date: November 8, 2024
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      # ratio: "16:9"
      # highlightLines: true
      # highlightStyle: solarized-light
      countIncrementalSlides: false
---

```{r child = "../setup.Rmd"}
```

```{css, echo = FALSE}
.tiny .remark-code { font-size: 70%; }
.small .remark-code { font-size: 80%; }
.tiny { font-size: 60%; }
.small { font-size: 80%; }
```


```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr); library(ggplot2)
# library(knitr)
```

## Announcements : Midterm 2

- Midterm next Friday 11/15
  - Will cover material from after Midterm 1 until today's lecture
  - Monday: holiday
  - Wednesday: review (OR)
  - Thursday: no lab, XHT OH 12-1pm on Zoom
  - Same rules apply: closed-book, no computers or calculators, no make-up exams

- These formulas will be provided: 
  - **Bayes' theorem**: $P(A \mid B) =\frac{P(B \mid A)P(A)}{P(B)}$.
  - **Probability mass functions**:  
      - Binomial: $P(X=x)=\begin{pmatrix} n \\ x \end{pmatrix}p^x(1-p)^{n-x}$  
      - Poisson: $P(X = x) = \frac{\lambda^x e^{-\lambda}}{x!}$, $\lambda > 0$


---
## Today
- Common probability distributions

  - Poisson distribution 


---

## Random samples from binomial distribution

- Use `rbinom()` to get 5000 draws from the population 

- In R:
.small[
```{r}
set.seed(0) # so results are reproducible 
binomDraws <- rbinom(n = 5000, size = 3, prob = .2)
table(binomDraws)/5000
```
]

---

## Frequency distribution for Binomial(3, .2)

```{r}
data.frame(binomDraws) %>%
  ggplot(aes(x = binomDraws)) + 
    geom_bar() +
    labs(x = "Number of Smokers",
         title = "5000 samples from Binomial(3, .2)")
```

---
## Varying the number of Bernoulli trials: 100 trials

.small[
```{r}
set.seed(0) # so results are reproducible 
binomDraws100 <- rbinom(n = 5000, size = 100, prob = .2)
data.frame(binomDraws100) %>%
  ggplot(aes(x = binomDraws100)) + 
    geom_bar() +
    labs(x = "Number of Smokers",
         title = "5000 samples from Binomial(100, .2)")
```
]
---
## Varying the number of Bernoulli trials: 500 trials

.small[
```{r}
set.seed(0) # so results are reproducible 
binomDraws500 <- rbinom(n = 5000, size = 500, prob = .2)
data.frame(binomDraws500) %>%
  ggplot(aes(x = binomDraws500)) + 
    geom_bar() +
    labs(x = "Number of Smokers",
         title = "5000 samples from Binomial(500, .2)")
```
]
---
## Frequency distribution varying number of Bernoulli trials

.small[
.pull-left[
```{r eval = FALSE}
data.frame(binomDraws) %>%
  bind_cols(size = 3) %>%
  bind_rows(
    data.frame(binomDraws100) %>%
      rename(binomDraws = binomDraws100) %>%
      bind_cols(size = 100)
  ) %>%
  bind_rows(
    data.frame(binomDraws500) %>%
      rename(binomDraws = binomDraws500) %>%
      bind_cols(size = 500)
  ) %>%
  ggplot(aes(x = binomDraws, 
             fill = as.factor(size))) +
    geom_histogram(binwidth = 1, position = "identity", alpha = .7) + 
    labs(
      x = "Number of smokers",
      y = "Frequency",
      title = "5000 samples each from Binomial(3, .2), Binomial(100, .2), Binomial(500, .2)",
      fill = "Size"
    )
```
]
]
.pull-right[
```{r echo = FALSE, out.width = "100%"}
data.frame(binomDraws) %>%
  bind_cols(size = 3) %>%
  bind_rows(
    data.frame(binomDraws100) %>%
      rename(binomDraws = binomDraws100) %>%
      bind_cols(size = 100)
  ) %>%
  bind_rows(
    data.frame(binomDraws500) %>%
      rename(binomDraws = binomDraws500) %>%
      bind_cols(size = 500)
  ) %>%
  ggplot(aes(x = binomDraws, 
                    fill = as.factor(size))) + 
    geom_histogram(binwidth = 1, position = "identity", alpha = .7) + 
    labs(
      x = "Number of smokers",
      y = "Frequency",
      title = "5000 samples each from \n Binomial(3, .2), Binomial(100, .2), Binomial(500, .2)",
      fill = "Size"
    )
```
]

---
## Frequency distribution varying probability of success

.small[
```{r }
set.seed(0) # so results are reproducible 
binomP.2 <- rbinom(n = 5000, size = 100, prob = .2)
binomP.5 <- rbinom(n = 5000, size = 100, prob = .5)
binomP.7 <- rbinom(n = 5000, size = 100, prob = .7)
```
]
```{r echo = FALSE, out.width = "60%"}
data.frame(binomP.2) %>%
  rename(outcome = binomP.2) %>%
  bind_cols(prob = .2) %>%
  bind_rows(
    data.frame(binomP.5) %>%
      rename(outcome = binomP.5) %>%
      bind_cols(prob = .5)
  ) %>%
  bind_rows(
    data.frame(binomP.7) %>%
      rename(outcome = binomP.7) %>%
      bind_cols(prob = .7)
  ) %>%
  ggplot(aes(x = outcome, 
                    fill = as.factor(prob))) +
    geom_histogram(binwidth = 1, position = "identity", alpha = .7) + 
    labs(
      x = "Number of smokers",
      y = "Frequency",
      title = "5000 samples each from \nBinomial(100, .2), Binomial(100, .5), Binomial(100, .7)",
      fill = "Prob"
    )
```



---
## Poisson distribution

- Useful for estimating the **number of events in a large population over a unit of time**. 

- For example:
  - The number of people having heart attacks in New York City every year
  - The number of accidents occurring at an intersection per hour
  - The number of typos in every 100 pages of a book 
  
- It is named after French mathematician Siméon Denis Poisson

---
## Poisson distribution
- E.g.: Number of people having heart attacks in New York City every year

- **Key ingredients**
  - **Fixed interval** of time or space
  
  - Events happen with a **known average rate**, independently of time since the last event ("memoryless" property)
      - One person having a heart attack does not change the probability of another person having a heart attack, hence the timing of the next heart attack 

- The parameter that defines a Poisson distributed random variable is the **rate** $\lambda$, where $\lambda > 0$

  - Rate = **average number of occurrences per unit of time**

- Often used to model rare events

---
## Probability mass function, mean and variance

- $P(X = x) = \frac{\lambda^x e^{-\lambda}}{x!}$, defined over non-negative integer values of $x$

  - Recall: $n! = n(n - 1)(n - 2)\cdots (1)$. 

- No upper limit, i.e., $x$ can take very large non-negative integer values 

- $E(X) = \lambda$

- $Var(X) = \lambda$

---
## Poisson probabilities  
- $P(X = x) = \frac{\lambda^x e^{-\lambda}}{x!}$ lets us calculate probabilities of taking a certain value

- For $x = 2$ and $\lambda = 3$, we have 

$$
\begin{aligned}
P(X = 2) &= \frac{3^2 e^{-3}}{2!}  = \frac{9(e^{-3})}{2(1)}  = 0.2240418
\end{aligned}
$$
- In R:

```{r}
dpois(x = 2, lambda = 3)
```

- For large values of $x$, the probability is very small because of the large denominator

```{r}
dpois(x = 10, lambda = 3)
```

---
## Probability distribution
- In the same manner, we can derive the entire probability distribution

.tiny[
.pull-left[
```{r}
dpois(x = 0:10, lambda = 3)
```

```{r eval = FALSE}
data.frame(x = 0:10, y = dpois(0:10, lambda = 3)) %>%
  ggplot(aes(x = x, y = y)) + 
    geom_bar(stat = "identity") +
  labs(title = "Probability distribution of Poisson(3)",
       y = "P(X = x)")
```
]
]
.pull-right[
```{r echo = FALSE, out.width = "100%"}
data.frame(x = 0:10, y = dpois(0:10, lambda = 3)) %>%
  ggplot(aes(x = x, y = y)) + 
    geom_bar(stat = "identity") +
  labs(title = "Probability distribution of Poisson(3)",
       y = "P(X = x)")
```
]

---
## Probability distribution varying lambda

.small[
```{r}
data.frame(x = 0:30, y = dpois(0:30, lambda = 3), lambda = 3) %>%
  bind_rows(data.frame(x = 0:30, y = dpois(0:30, lambda = 10), lambda = 10)) %>%
  bind_rows(data.frame(x = 0:30, y = dpois(0:30, lambda = 20), lambda = 20)) %>%
    ggplot(aes(x = x, y = y, fill = as.factor(lambda))) + 
      geom_bar(stat = "identity", 
               position = "identity", 
               alpha = .5) +
    labs(title = "Probability distribution of \nPoisson(3), Poisson(10), Poisson(20)",
         y = "P(X = x)",
         fill = "Lambda")
```
]

---
## Sampling from Poisson distribution in R
- Simulate random draws using the `rpois()` function

- `rpois()` has the arguments 
  - `n`, the number of draws from the distribution 
  - `lambda`, the mean

```{r}
set.seed(0) # so results are reproducible 
inputLambda <- 3
poissonDraws <- rpois(n = 100, lambda = inputLambda)
poissonDraws
```

---
## Frequency distribution varying lambda

.small[
```{r }
set.seed(0) # so results are reproducible 
poissonL3 <- rpois(n = 5000, lambda = 3)
poissonL10 <- rpois(n = 5000, lambda = 10)
poissonL20 <- rpois(n = 5000, lambda = 20)
```
]

.tiny[
.pull-left[
```{r eval = FALSE}
data.frame(poissonL3) %>%
  rename(outcome = poissonL3) %>%
  bind_cols(lambda = 3) %>%
  bind_rows(
    data.frame(poissonL10) %>%
      rename(outcome = poissonL10) %>%
      bind_cols(lambda = 10)
  ) %>%
  bind_rows(
    data.frame(poissonL20) %>%
      rename(outcome = poissonL20) %>%
      bind_cols(lambda = 20)
  ) %>%
  ggplot(aes(x = outcome, 
                    fill = as.factor(lambda))) +
    geom_histogram(binwidth = 1, position = "identity", alpha = .7) + 
    labs(
      x = "Number of occurrences",
      y = "Frequency",
      title = "5000 samples each from \nPoisson(3), Poisson(10), Poisson(20)",
      fill = "Lambda"
    )
```
]
]
.pull-right[
```{r echo = FALSE, out.width = "100%"}
data.frame(poissonL3) %>%
  rename(outcome = poissonL3) %>%
  bind_cols(lambda = 3) %>%
  bind_rows(
    data.frame(poissonL10) %>%
      rename(outcome = poissonL10) %>%
      bind_cols(lambda = 10)
  ) %>%
  bind_rows(
    data.frame(poissonL20) %>%
      rename(outcome = poissonL20) %>%
      bind_cols(lambda = 20)
  ) %>%
  ggplot(aes(x = outcome, 
                    fill = as.factor(lambda))) +
    geom_histogram(binwidth = 1, position = "identity", alpha = .7) + 
    labs(
      x = "Number of occurrences",
      y = "Frequency",
      title = "5000 samples each from \nPoisson(3), Poisson(10), Poisson(20)",
      fill = "Lambda"
    )
```
]



---
## Exercises
An insurance agency determines that 70% of individuals do not exceed their deductible. 

- Suppose the insurance agency is considering a random sample of four individuals they insure. What is the probability that exactly one of them will exceed the deductible?

--

- What is the probability that 3 of 8 randomly selected individuals will have exceeded the insurance deductible, i.e., that 5 of 8 will not exceed the deductible?

---
## Exercises
A very skilled court stenographer makes one typographical error (typo) per hour on average.

- What probability distribution is most appropriate for calculating the probability of a given number of typos this stenographer makes in an hour?

- What are the mean and the standard deviation of the number of typos this stenographer makes?

- Would it be considered unusual if this stenographer made 4 or more typos in a given hour?

- Calculate the probability that this stenographer makes at most 2 typos in a given hour.


---
## Summary

- Common probability distributions: Poisson 

  - Theoretical properties: probability density function, parameters, mean and variance, effect of varying parameters
  
  - R functions, e.g.:
  
      - `dpois()` for densities 
      - `ppois()` for $P(X\leq x)$
      - `rpois()` for random sample
    

