---
title: "Binomial and Poisson Distribution"
subtitle: "<br><br> STA35A: Statistical Data Science 1"
author: "Xiao Hui Tai"
date: November 8, 2023
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      # ratio: "16:9"
      # highlightLines: true
      # highlightStyle: solarized-light
      countIncrementalSlides: false
---


```{r child = "../setup.Rmd"}
```

```{css, echo = FALSE}
.tiny .remark-code { font-size: 70%; }
.small .remark-code { font-size: 80%; }
.tiny { font-size: 60%; }
.small { font-size: 80%; }
```


```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
```


## Reminders/announcements

- Friday is a holiday! 
  - I will post homework by Friday, due Thursday 11/16 at 9pm
  
- Midterm next Friday 11/17
  - Will cover material from 10/23 (after Midterm 1) until next Monday 11/13
  - Wednesday: review (JH)
  - Thursday: no lab, XHT OH 2:30-3:30pm
  - Same rules apply: closed-book, no computers or calculators, no make-up exams

---

## Recap
--

- Common probability distributions

  - Bernoulli 
      - Theoretical properties: probability mass function, parameters, mean and variance 
      - Sampling and law of large numbers; effect of changing parameters 
  
  - Binomial

---
## Today
- Common probability distributions

  - Binomial 
      - Theoretical properties: probability mass function, parameters, mean and variance 
      - More on calculating binomial probabilities
      - Sampling and law of large numbers; effect of changing parameters 
  
  - Poisson
  
---

## Binomial random variable

- **Binomial distribution**: probability of $X$ "successes" from a sequence of $n$ independent Bernoulli trials (size $n$)

  - Binomial(n, p)

- E-cigarette example: each student would represent an independent Bernoulli trial 

  - 1 draw from the binomial distribution is made of 3 independent draws from the Bernoulli distribution 

- **Assumptions**:

  - Fixed number $n$ of Bernoulli trials
  
  - Outcomes of the $n$ trials are **independent**
  
  - Probability of success $p$ is the same for each trial

---

## Binomial distribution

- **Probability mass function**: $P(X=x)=\begin{pmatrix} n \\ x \end{pmatrix}p^x(1-p)^{n-x}$ 

- Recall Bernoulli: $P(Y=y)=p^y(1-p)^{1-y}$

- Second part: multiplying the right combination of $p$ and $1-p$ 

  - Total of $n$ terms 

  - For example, for 3 e-cig smokers, $X=3$, $p^x(1-p)^{n-x}=0.2^3(0.8)^0=0.008$

- First part: $\begin{pmatrix}n \\x \end{pmatrix}$: all the possible ways to get the outcome $x$


---

## Combinations

- $\begin{pmatrix} n \\ x \end{pmatrix}$, is said in words "n choose x" 

- This is called a **combination**.

- In this particular setting, it is also called the *binomial coefficient*. 

- Represents the number of ways to pick $x$ subjects from a larger group of $n$ and is defined as $$\begin{pmatrix} n \\ x \end{pmatrix} = \frac{n!}{x!(n-x)!}.$$

- n! ("n factorial"): $n!=n(n-1)(n-2)\cdots (1)$. 

  - $3!=3(2)(1)=6$
  - $4!=4(3)(2)(1)=24$, and so forth. 
  - We define $0!=1$.

---

## Combinations
- In R, `choose(n, k)`

- Factorials are `factorial(x)`, e.g., 
```{r}
factorial(3)
```

- How many ways can we pick 3 subjects from a group of 3? Using the binomial coefficient, we see it is $\begin{pmatrix}3 \\ 3 \end{pmatrix}=\frac{3!}{3!0!}=\frac{3(2)(1)}{3(2)(1)1}=1$. In R,
```{r}
choose(3, 3)
```

---

## Combinations

Going back to the table: 3 ways to pick 2 from a group of 3 students.  

.small[

| $Y_1$ | $Y_2$ | $Y_3$ | $X$ | $P(X)$ |
|:----:|:----:|:----:|:----:|:----:|
| 0 | 0 | 0 | 0 | 0.8(0.8)(0.8)=0.512 |
| 1 | 0 | 0 | 1 | 0.2(0.8)(0.8)=0.128 |
| 0 | 1 | 0 | 1 | 0.8(0.2)(0.8)=0.128 |
| 0 | 0 | 1 | 1 | 0.8(0.8)(0.2)=0.128 |
| 1 | 1 | 0 | 2 | 0.2(0.2)(0.8)=0.032 |
| 1 | 0 | 1 | 2 | 0.2(0.8)(0.2)=0.032 |
| 0 | 1 | 1 | 2 | 0.8(0.2)(0.2)=0.032 |
| 1 | 1 | 1 | 3 | 0.2(0.2)(0.2)=0.008 |

]

We can also calculate $\begin{pmatrix}3 \\ 2 \end{pmatrix}=\frac{3!}{2!1!}=\frac{3(2)(1)}{(2)(1)(1)}=3$.

---

## Binomial probabilities

Putting the ingredients together, we have 

$$
\begin{aligned}
P(X = 2) &=  \begin{pmatrix} n \\ x \end{pmatrix}p^x(1-p)^{n-x}\\
&= \begin{pmatrix} 3 \\ 2 \end{pmatrix}p^2(1-p)^{3-2}\\
&=3(0.2)^2(0.8)^{3-2}\\
&=3(.032) \\
&=.096
\end{aligned}
$$
This is the same as we saw in the table. 

---

## Binomial probabilities 

- We saw how to calculate $P(X = 2)$. How about $P(X = 3)$?

$$
\begin{aligned}
P(X = x) &=\begin{pmatrix} n \\ x \end{pmatrix} p^x(1-p)^{n-x} \\
&= \frac{n!}{x!(n-x)!}p^x(1-p)^{n-x} 
\end{aligned}
$$

--

- Probability of getting 3 recent e-cig smokers in our sample of 3 students,

$$\begin{aligned}
P(X = 3)& = \begin{pmatrix} 3 \\ 3 \end{pmatrix}0.2^3(1-0.2)^{3-3}  \\
& = \frac{3!}{3!(3-3)!}0.2^3(1-0.2)^{3-3} \\
 & = \frac{3(2)(1)}{3(2)(1)1}0.2^3(0.8)^{0}=0.2^3 \\
 & = 0.008.
 \end{aligned}$$

- Derive the probability distribution

---
## Binomial probabilites in R

- In R, we can use `dbinom()`

- From the help page: `dbinom()` has the arguments `x, size, prob, log = FALSE`
  - `x` are the values of $x$ that we want probabilities for; here it is 0, 1, 2 and 3
  - `size` is the number of Bernoulli trials; here it is 3
  - `prob` is the probability of success; here it is .2

.pull-left[
```{r}
dbinom(x = 0:3, size = 3, 
       prob = .2)
```
]

.pull-right[

| | | | | |
|:--:|:--:|:--:|:--:|:---:|
| $X$ | 0 | 1 | 2 | 3|
| $P(X=x)$ | 0.512 | 0.384 | 0.096 | 0.008 |
 
 ]


---
## Bernoulli vs. binomial random variable

- Recall: The **binomial distribution** gives us the probability of $X$ "successes" from a sequence of $n$ independent Bernoulli trials. 

- In our example, each student would be an independent Bernoulli trial (either an e-cig smoker, or not).

  - **Bernoulli**: whether or not a student is a recent e-cigarette smoker 

  - **Binomial**: number of students, out of 3, who are recent e-cigarette smokers

  - One observation of this binomial random variable is composed of 3 Bernoulli trials ("size 3")

- Bernoulli distribution = special case of the binomial distribution, with 1 "trial"

---
## Mean and variance of binomial random variable

For a binomial random variable with number of trials $n$ and probability of success $p$, $E(X) = np$ and $Var(X) = n(p)(1-p)$

- How to get this? 

--

- From Bernoulli!

- In our example, $p = .2$, $n = 3$, so $E(X) = .6$ and $Var(X) = 3*.2*.8 = .48$

--

- The Bernoulli distribution is a special case of the binomial distribution, with $n = 1$. Recall: $E(X) = p$ and $Var(X) = p(1-p)$.

---

## Case Study: E-Cigarettes

- The [CDC reports](https://www.cdc.gov/tobacco/data_statistics/fact_sheets/youth_data/tobacco_use/index.htm) that about 20% of high school students have smoked e-cigarettes in the past 30 days

- $P(Y=1)=P(Smoker)=p=0.2$ and $P(Y=0)=0.8$

- Previously: 3 Bernoulli trials
  

---
## More examples of calculating binomial probabilities 
- What if we are interested in the probability that there is at least 1 student, out of 500, that has recent e-cigarette use? 

- $X =$? 

- What is the probability of interest? 

- $n =$? $p =$?

---
## More examples of calculating binomial probabilities 
- What if we are interested in the probability that there are more than 2 students, out of 500, who have recent e-cigarette use? 

--

Binomial distribution: $P(X = x)=\begin{pmatrix} n \\ x \end{pmatrix}p^x(1-p)^{n-x}$

- Quick trick: $P(X \geq 1) = 1 - P(X=0)$ (either no students with recent e-cigarette use, or one or more students with recent e-cigarette use)

- Similarly, $P(X > 2)=1-P(X=0)-P(X=1)-P(X=2)$ 


---
## More examples of calculating binomial probabilities 

Binomial distribution: $P(X = x)=\begin{pmatrix} n \\ x \end{pmatrix}p^x(1-p)^{n-x}$

So we have

.small[
$$
\begin{aligned}
P(X \geq 1) & =1-P(X = 0) \\
&=1-\begin{pmatrix}500 \\ 0 \end{pmatrix} 0.2^0 (0.8)^{500-0} \\
&=1-0.8^{500} \\
&= 1\\
 P(X>2) & = 1-P(X=0)-P(X=1)-P(X=2) \\ 
 & = 1 - 0.8^{500} - \begin{pmatrix} 500 \\ 1 \end{pmatrix}0.2^1(0.8)^{500-1} - \begin{pmatrix} 500 \\ 2 \end{pmatrix}0.2^2(0.8)^{500-2} \\
 & = 1 - 0.8^{500} - 500(0.2)(0.8)^{500-1} - \frac{500!}{2!(500-2)!}0.2^2(0.8)^{500-2} \\
 & = 1 - 0.8^{500} - 500(0.2)(0.8)^{500-1} - \frac{500(500-1)}{2}0.2^2(0.8)^{500-2} \\
 & = 1
\end{aligned}
$$
]

---
## In R

- Given the large number of students (500), it is almost certain that at least 1 or 2 would have recent e-cigarette use. 

- We might be interested instead in $P(X \geq 100)$, the probability that at least four-fifths of the students have recent e-cigarette use

- Luckily, we can do this in R

```{r}
sum(dbinom(x = 100:500, size = 500, prob = .2))
```

- Also, `pbinom()`: gives us $P(X \leq x)$ directly
```{r}
1 - pbinom(q = 99, size = 500, prob = .2)
```

---
## Sampling from a binomial distribution in R

- Recall draws from the Bernoulli distribution:
  
  - For any value of $0 \leq p \leq 1$ (including .5), we can use the `rbinom()` function

  - `rbinom()` has the arguments `n, size, prob`, where `n` is the number of draws, and `prob` is the probability of success.
  
  - `size` is the "number of trials (zero or more)" -- for the Bernoulli distribution, we use `size = 1`. 
  
---
## Sampling from a binomial distribution in R

- `size` = number of Bernoulli trials; in the 3-student e-cigarette example, `size = 3`

- **NOTE**: we use $n$ to denote the number of Bernoulli trials; in R this is the `size` argument

Description | R
----|--
Number of draws/samples from Binomial | `n`
Number of Bernoulli trials, size $n$ | `size`
Probability of success, $p$ | `prob`

- Bernoulli distribution: `size = 1`

---
## Sampling from a binomial distribution in R

Example: 100 draws from the binomial distribution; each draw is made of 3 Bernoulli trials (size 3); probability of success = .2

```{r}
set.seed(0) # so results are reproducible 
inputP <- .2
binomDraws <- rbinom(n = 100, size = 3, prob = inputP)
binomDraws
```


---
## Sample mean  

- Recall: for a binomial random variable with number of trials $n$ and probability of success $p$, $E(X) = np$ and $Var(X) = n(p)(1-p)$

- In our example, $p = .2$, $n = 3$, so $E(X) = .6$ and $Var(X) = 3*.2*.8 = .48$

- We can calculate the sample mean from our sample of 100:
```{r}
mean(binomDraws)
```

- What should this converge to? 

---
## Law of Large Numbers again
As the sample size grows, the sample mean gets closer to the expected value, or population mean

.small[
```{r}
set.seed(0) # so results are reproducible 
binomDraws <- rbinom(n = 5000, size = 3, prob = .2)
myMeans <- data.frame(sampleSize = 1:5000, myMean = NA)
meanFun <- function(inputSampSize, outcomes) {
  return(mean(outcomes[1:inputSampSize]))
}
myMeans$myMean <- sapply(myMeans$sampleSize, meanFun, binomDraws)
head(myMeans)
tail(myMeans)
```
]
---
## Law of Large Numbers again 
.small[
```{r}
myMeans %>%
  ggplot(aes(x = sampleSize, y = myMean)) +
  geom_line() +
  labs(x = "Sample size",
       y = "Sample mean",
       title = "Illustration of law of large numbers for Binomial distribution\nn = 3, p = .2")
```
]


---

## Frequency distribution vs. probability distribution 

- Use `rbinom()` to get 5000 draws from the population 

- In R:
.small[
```{r}
set.seed(0) # so results are reproducible 
binomDraws <- rbinom(n = 5000, size = 3, prob = .2)
table(binomDraws)/5000
```
]

- Compare with the theoretical probabilities:
.small[
```{r}
dbinom(x = 0:3, size = 3, prob = .2)
```
]
---

## Frequency distribution of e-cigarette smokers 

```{r}
data.frame(binomDraws) %>%
  ggplot(aes(x = binomDraws)) + 
    geom_bar() +
    labs(x = "Number of Smokers",
         title = "5000 samples from Binomial(3, .2)")
```

---
## Varying the number of Bernoulli trials: 100 trials

.small[
```{r}
set.seed(0) # so results are reproducible 
binomDraws100 <- rbinom(n = 5000, size = 100, prob = .2)
data.frame(binomDraws100) %>%
  ggplot(aes(x = binomDraws100)) + 
    geom_bar() +
    labs(x = "Number of Smokers",
         title = "5000 samples from Binomial(100, .2)")
```
]
---
## Varying the number of Bernoulli trials: 500 trials

.small[
```{r}
set.seed(0) # so results are reproducible 
binomDraws500 <- rbinom(n = 5000, size = 500, prob = .2)
data.frame(binomDraws500) %>%
  ggplot(aes(x = binomDraws500)) + 
    geom_bar() +
    labs(x = "Number of Smokers",
         title = "5000 samples from Binomial(500, .2)")
```
]
---
## Frequency distribution of e-cigarette smokers varying number of Bernoulli trials

.small[
.pull-left[
```{r eval = FALSE}
data.frame(binomDraws) %>%
  bind_cols(size = 3) %>%
  bind_rows(
    data.frame(binomDraws100) %>%
      rename(binomDraws = binomDraws100) %>%
      bind_cols(size = 100)
  ) %>%
  bind_rows(
    data.frame(binomDraws500) %>%
      rename(binomDraws = binomDraws500) %>%
      bind_cols(size = 500)
  ) %>%
  ggplot(aes(x = binomDraws, 
             fill = as.factor(size))) +
    geom_histogram(binwidth = 1, position = "identity", alpha = .7) + 
    labs(
      x = "Number of smokers",
      y = "Frequency",
      title = "5000 samples each from Binomial(3, .2), Binomial(100, .2), Binomial(500, .2)",
      fill = "Size"
    )
```
]
]
.pull-right[
```{r echo = FALSE, out.width = "100%"}
data.frame(binomDraws) %>%
  bind_cols(size = 3) %>%
  bind_rows(
    data.frame(binomDraws100) %>%
      rename(binomDraws = binomDraws100) %>%
      bind_cols(size = 100)
  ) %>%
  bind_rows(
    data.frame(binomDraws500) %>%
      rename(binomDraws = binomDraws500) %>%
      bind_cols(size = 500)
  ) %>%
  ggplot(aes(x = binomDraws, 
                    fill = as.factor(size))) + 
    geom_histogram(binwidth = 1, position = "identity", alpha = .7) + 
    labs(
      x = "Number of smokers",
      y = "Frequency",
      title = "5000 samples each from \n Binomial(3, .2), Binomial(100, .2), Binomial(500, .2)",
      fill = "Size"
    )
```
]

---
## Frequency distribution of e-cigarette smokers varying probability of success

.small[
```{r }
set.seed(0) # so results are reproducible 
binomP.2 <- rbinom(n = 5000, size = 100, prob = .2)
binomP.5 <- rbinom(n = 5000, size = 100, prob = .5)
binomP.7 <- rbinom(n = 5000, size = 100, prob = .7)
```
]
```{r echo = FALSE, out.width = "60%"}
data.frame(binomP.2) %>%
  rename(outcome = binomP.2) %>%
  bind_cols(prob = .2) %>%
  bind_rows(
    data.frame(binomP.5) %>%
      rename(outcome = binomP.5) %>%
      bind_cols(prob = .5)
  ) %>%
  bind_rows(
    data.frame(binomP.7) %>%
      rename(outcome = binomP.7) %>%
      bind_cols(prob = .7)
  ) %>%
  ggplot(aes(x = outcome, 
                    fill = as.factor(prob))) +
    geom_histogram(binwidth = 1, position = "identity", alpha = .7) + 
    labs(
      x = "Number of smokers",
      y = "Frequency",
      title = "5000 samples each from \nBinomial(100, .2), Binomial(100, .5), Binomial(100, .7)",
      fill = "Prob"
    )
```



---
## Poisson distribution

- Useful for estimating the **number of events in a large population over a unit of time**. 

- For example:
  - The number of people having heart attacks in New York City every year
  - The number of accidents occurring at an intersection per hour
  - The number of typos in every 100 pages of a book 
  
- It is named after French mathematician Siméon Denis Poisson

---
## Poisson distribution
- E.g.: Number of people having heart attacks in New York City every year

- **Key ingredients**
  - **Fixed interval** of time or space
  
  - Events happen with a **known average rate**, independently of time since the last event ("memoryless" property)
      - One person having a heart attack does not change the probability of another person having a heart attack, hence the timing of the next heart attack 

- The parameter that defines a Poisson distributed random variable is the **rate** $\lambda$, where $\lambda > 0$

  - Rate = **average number of occurrences per unit of time**

- Often used to model rare events

---
## Probability mass function, mean and variance

- $P(X = x) = \frac{\lambda^x e^{-\lambda}}{x!}$, defined over non-negative integer values of $x$

  - Recall: $n! = n(n - 1)(n - 2)\cdots (1)$. 

- No upper limit, i.e., $x$ can take very large non-negative integer values 

- $E(X) = \lambda$

- $Var(X) = \lambda$

---
## Poisson probabilities  
- $P(X = x) = \frac{\lambda^x e^{-\lambda}}{x!}$ lets us calculate probabilities of taking a certain value

- For $x = 2$ and $\lambda = 3$, we have 

$$
\begin{aligned}
P(X = 2) &= \frac{3^2 e^{-3}}{2!}  = \frac{9(e^{-3})}{2(1)}  = 0.2240418
\end{aligned}
$$
- In R:

```{r}
dpois(x = 2, lambda = 3)
```

- For large values of $x$, the probability is very small because of the large denominator

```{r}
dpois(x = 10, lambda = 3)
```

---
## Exercises
An insurance agency determines that 70% of individuals do not exceed their deductible. 

- Suppose the insurance agency is considering a random sample of four individuals they insure. What is the probability that exactly one of them will exceed the deductible?

--

- What is the probability that 3 of 8 randomly selected individuals will have exceeded the insurance deductible, i.e., that 5 of 8 will not exceed the deductible?

---
## Exercises
A very skilled court stenographer makes one typographical error (typo) per hour on average.

- What probability distribution is most appropriate for calculating the probability of a given number of typos this stenographer makes in an hour?

- What are the mean and the standard deviation of the number of typos this stenographer makes?

- Would it be considered unusual if this stenographer made 4 or more typos in a given hour?

- Calculate the probability that this stenographer makes at most 2 typos in a given hour.


---
## Summary

--

- Common probability distributions: Binomial and Poisson 

  - Theoretical properties: probability mass function, parameters, mean and variance, effect of varying parameters
  
  - Sampling and law of large numbers; effect of changing parameters 

  - R functions:
  
      - `d____()`, e.g., `dbinom()`: for densities (more accurately, for discrete random variables these are probability mass functions, $P(X = x)$)
      - `p____()`, e.g., `pbinom()`: for $P(X\leq x)$
      - `r____()`, e.g., `rbinom()`: for random sample
    
